---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### Loading and preprocessing the data

We begin by loading libraries, clearing the workspace, and unzipping the `activity.zip` dataset, which includes the comma-separated `activity.csv` file, which contains the raw dataset for this analysis.
```{r setup, echo=T, message=F, warning=F}
library(dplyr); library(mice); library(lattice); library(moments); library(lambda.tools)
knitr::opts_chunk$set(cache=TRUE)
# Set working directory (edit this path as desired), fetch data to it
dataDirName <- "C:/Users/tom/Documents/DataScience/course 5/wk2/RepData_PeerAssessment1"
setwd(dataDirName)  # make it working directory
# skip the unzip if activity.csv already exists; otherwise, unzip it to current working directory
if( !file.exists("./activity.csv") ) unzip("./activity.zip", overwrite = T, exdir = getwd())
activity <- read.csv("activity.csv", header = T, sep = ",")
```
The activity dataset consists of `r dim(activity)[1]` observations across `r dim(activity)[2]` variables.  We observe some missing values in the `steps` field, denoted as NAs in the summary below.  These will be discussed later in the report.  For now we will ignore the missing values.
```{r summary, echo=T}
str(activity)
```

### Mean and median total number of steps taken per day

The **mean** and **median** total number of steps taken per day are reported below.  In addition, a histogram representing the frequency distribution of the total number of steps taken each day is presented.  The **mean** and **median** values (dark red line, dashed blue line, respectively) are shown as vertical intercept lines on the plot.  We observe that the **mean** is slightly less than the **median**, so this is a slightly left-skewed distribution.
```{r stats, fig.align='center', fig.width=9, fig.height=5, echo=T}
sumSteps <- activity %>% group_by(date) %>% summarize(sum=sum(steps, na.rm=TRUE))
cat("The data spans", dim(sumSteps)[1], "days.",
     "\nStatistics, steps taken per day:\tMean:", (mn <- round(mean(sumSteps$sum),0)),
	"\tMedian:", (md <- round(median(sumSteps$sum),0)))
hst <- function() {
     hist(sumSteps$sum/1000, breaks=20, main='Total Steps per Day, 1,000s', xlab='',
     	ylab='Percent of Total', xlim=c(0,25), col='springgreen', axes=F)
     axis(2)
     axis(1, at=seq(0,ceiling(max(sumSteps$sum)/1000), by=2),
     	labels=seq(0,ceiling(max(sumSteps$sum)/1000), by=2))
     abline(v=md/1000, col='royalblue4', lwd=2, lty=2)  # dashed blue line to indicate median
     abline(v=mn/1000, col='firebrick', lwd=2)  # dark red line to indicate mean
     text(mn/1000-1.5, 10.25, "mean >>", col = 'firebrick')
     text(md/1000+2, 10.25, "<< median", col = 'royalblue4')
}
hst()  # display it
cat("Skewness:", round(skewness(sumSteps$sum),2))
```
**The mean and median total steps per day values for the original dataset are `r format(mn, nsmall=0)` and `r format(md, nsmall=0)`, respectively.**

```{r stats II, fig.align='center', fig.width=9, fig.height=5, include=F, echo=T}
pdf(file="./figure/histogram - Total Steps per Day.pdf", width=7, height=5)
hst(); dev.off()  # pdf it
```
A negative skewness value close to zero indicates a slight left-skew;  symmetric distributions have zero skew.  We note the very large proportion at the far left of the plot.

### What is the average daily activity pattern?

The total number of steps taken each day during the period `r paste(gsub("2012-", "", c(as.character(sumSteps$date[1]), as.character(sumSteps$date[length(sumSteps$date)]))), collapse=" to ")` is presented in the barplot below.
```{r barplot_daily_activity, fig.align='center', fig.width=9, fig.height=5, echo=T}
{par(las=2) # x-axis label orientation perpendicular
barplot(height=sumSteps$sum/1000,
        main="Daily Activity Pattern in 2012", xlab="", ylab="Total Steps (thousands)",
        names.arg = sub("2012-", "", sumSteps$date), col='springgreen', cex.names = .8,
        cex.axis = 1.5, cex.lab=1.3)
par(las=0)}
```
We will come back to the daily activity patterns later in the report, specifically to determine if there are differences between weekday and weekend-day steps activity.

This dataset contains `r length(unique(activity$interval))` intervals per day, each five minutes in duration.  We are interested in the daily activity pattern, specifically the average number of steps taken in each five-minute interval, averaged across all days. This pattern is presented on the plot below.
```{r avg activity per interval, fig.align='center', fig.width=9, fig.height=5, echo=T}
# calculate average number of steps per interval across all dates
avgSteps <- activity %>% group_by(interval) %>% summarize(average=mean(steps, na.rm=TRUE))
# which interval across all dates contains highest number of steps?
mx_idx<-with(avgSteps, which(average==max(average))) # interval index with max average
mx_val<-with(avgSteps, interval[which(average==max(average))]) # max interval average value
plot(y=avgSteps$average, x=avgSteps$interval, main="Average steps time series analysis",
     xlab="Interval", ylab="Average Steps", col='firebrick', type = "l")
abline(v=mx_val, lwd=2.5, lty=2, col = 'royalblue4')
abline(v=1200, lwd=1, lty=3)
text(mx_val-175, 205, paste0(mx_val, " is max >>"), col = 'royalblue4')
text(1300, 205, "~ noon", col = 'black')
```

We observe a number of activity spikes during the day, the **largest of which at interval `r mx_val`**, which likely corresponds to a activity just after 8:00AM.  The maximum value (`r mx_val`) is highlighted on the chart as a dashed vertical line.  We also observe spikes in activity mid-day (dotted line), at mid-afternoon, and at about 4PM and 6PM.

### Imputing missing values

As noted earlier, there are a significant number of observations with missing values (coded as `NA`) in the `steps` field. These are a potential source of undesirable bias.  The analysis so far has ignored missing values, but we will explore one method for imputing missing values.

```{r NA summary I, echo=T}
cat("The steps variable includes", (s<-sum(is.na(activity$steps))), "NAs, or ",
	round(100*s/length(activity$steps),0), "% of observations")
```
We note a high proportion of missing values in `steps`, and they cluster throughout the observations (including at the beginning and end) in the `activity` dataset, as tabulated below.  We use the range_for() function to easily identify ranges for missing values.
```{r NA summary II, echo=T}
NAspan <- range_for(TRUE, is.na(activity$steps))
data.frame(`lower bound`=NAspan$min, `upper bound`=NAspan$max)
```

The author chose not to use simpler methods for imputing missing values, such as substituting column means or medians.  Instead, we will examine the MICE method to impute missing values - MICE is an acronym for Multivariate Imputation by Chained Equations.  The associated function is mice(), found in the `mice` library, which was loaded earlier.  The author experimented with the following mice() function variables: `m` (number of multiple imputations), `method` (imputation method), and `maxit` (number of iterations), specifically the value ranges m={5,7,8}, method={fastpmm, pmm}, maxit={10,20,30,50,80,100}.

Fit was evaluated with a densityplot, the author determined that the m=5, method=pmm, maxit=100 parameter values yielded suitable results.  That densityplot is included as a pdf in the /figure directory of this github listing.  On the plot, magenta denotes the imputed values - we observe that they cluster approximately around original, valid points (the blue line), with notable exceptions at very low values.

The `date` and `interval` column data is unchanged by this operation - the author checks this assumption below. 
```{r imputation I, echo=TRUE}
micedActivity <- mice(data = activity, m = 5, method = "pmm", maxit = 100, seed=2222, print=F)
```
```{r imputation II, echo=T}
newActivity <- complete(micedActivity, 1)  # extract the new dataframe
# show that `date` and `interval` variables are unchanged from original dataset
{if(sum(newActivity$date!=activity$date) == 0 ) cat("post-imputation:  date variable unaltered")
if(sum(newActivity$interval!=activity$interval) == 0 ) cat("\t\tinterval variable unaltered")}
```
```{r imputation IIa, fig.align='center', fig.width=7, fig.height=5, include=F, echo=T}
pdf(file="./figure/densityplot - Validation of imputed value fit.pdf", width=7, height=5)
densityplot(micedActivity, ylim=c(0,.01), main="Density plot, imputed and original values")
hst(); dev.off()  # pdf it
```
#### Calculate statistics for imputed dataset
```{r imputation III, fig.align='center', fig.width=9, fig.height=5, echo=T}
revSumSteps <- newActivity %>% group_by(date) %>% summarize(sum=sum(steps, na.rm=TRUE))
cat("Revised statistics, steps taken per day:\tMean:", (revMn <- round(mean(revSumSteps$sum),0)),
	"\tMedian:", (revMd <- round(median(revSumSteps$sum),0)))
hst <- function() {
     hist(revSumSteps$sum/1000, breaks=20, main='Revised Total Steps per Day, 1,000s', xlab='',
     	ylab='Percent of Total', xlim=c(0,25), col='springgreen', axes=F)
     axis(2)
     axis(1, at=seq(0,ceiling(max(revSumSteps$sum)/1000), by=2),
     	labels=seq(0,ceiling(max(revSumSteps$sum)/1000), by=2))
     abline(v=revMd/1000, col='royalblue4', lwd=2, lty=2)  # dashed blue line to indicate median
     abline(v=revMn/1000, col='firebrick', lwd=2)  # dark red line to indicate mean
}
hst()  # display it
```
```{r imputation IIII, fig.align='center', fig.width=9, fig.height=5, include=F, echo=T}
pdf(file="./figure/histogram - Revised Total Steps per Day.pdf", width=7, height=5)
hst(); dev.off()  # pdf it
```

Imputing missing `step` values has had an interesting impact on the distribution of the total steps taken per day, as shown above.  **The revised mean and median values are `r format(revMn, nsmall=0)` and `r format(revMd, nsmall=0)`, respectively.**  As before, the **revised mean** and **revised median** values are shown as vertical intercept lines on the histogram (dark red line, dashed blue line, respectively). We observe that **revised mean** and **revised median** values have both shift slightly to larger values after imputation, and they are nearly equal, which indicates a more symmetrical distribution as a result of imputation.

### Are there differences in activity patterns between weekdays and weekends?

Now we will explore any differences in activity patterns between weekdays and weekends as measured in the `step` variable.  We will add a new factor variable to the `activity` dataset to differentiate between weekend and weekday observations.

A panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days, is presented below.

```{r dayType, fig.align='center', fig.width=9, fig.height=5, echo=T}
library(lattice);library(ggplot2)
# create new factor variable `dayType`
dayType <- weekdays(as.Date(newActivity$date))
for(i in 1:length(dayType)) {
	if(dayType[i] == "Saturday" | dayType[i] == "Sunday") {
		dayType[i] <- "weekend"
	} else { dayType[i] <- "weekday"
	}
}
newActivity$dayType <- as.factor(dayType)
# calculate means by interval by dayType
dayTypeNewActivity <- 
     newActivity %>% group_by(interval, dayType) %>% summarize(meanSteps=round(mean(steps),2))
(g1 <- xyplot(meanSteps ~ interval | dayType, data = dayTypeNewActivity,
       type = "l", col="blue", grid=T, layout = c(1, 2),
       main="Activity comparisons, weekends vs. weekdays"))
g2 <- ggplot(data=dayTypeNewActivity, aes(x=meanSteps, fill=dayType)) +
     geom_histogram(color = "black", binwidth = 2) + ylim(0,30) + xlim(0,180) +
     facet_grid(. ~ dayType) + theme(legend.position="bottom") +
     ggtitle("Activity comparisons, weekdays vs. weekends")
```
```{r dayType II, include=F}
pdf(file="./figure/xyplot - Activity comparisons, weekdays vs. weekends.pdf", width=7, height=5)
print(g1); dev.off()
pdf(file="./figure/ggplot - Activity comparisons, weekdays vs. weekends.pdf", width=7, height=5)
print(g2); dev.off()
```

The weekday activity is represented by the plot above.  There is more sustained early-morning weekday activity for this individual as compared to the weekend case at top.  The weekend activity appears to increase during mid-day hours.  Overall, there appears to be more mid-day activity on weekends.  The author has also provided a ggplot() output as a pdf file.